The Unicode Cookbook \linebreak for Linguists

### 2016-08-18 book/chapters/implementation.tex
adding new chapter with R examples

\noindent The version available through CRAN is the most recent stable version.
To obtain the most recent bug-fixes and experimental additions, please use the
development versions, which is available on
GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
version can be easily installed using the github-install helper software from the
\texttt{devtools} package.

<<eval=FALSE>>=
# download and install helper software
install.packages('devtools') 
# install the qlcData package from GitHub
devtools::install_github('cysouw/qlcData')
# load the software, so it can be used 
library(qlcData) 
@

@@

\noindent Some of these graphemes are single codepoints, others are combinations
of two codepoints. By specifying the splitting separator as empty, it is
possible to split the string into Unicode codepoints.

@@

\noindent Some characters look identical, although they are encoded differently. 
Unicode offers different ways of normalization, which can be invoked here as 
well (see Section~\ref{pitfall-canonical-equivalence}).

@@

\noindent It is important to realize that for Unicode grapheme definitions, NFC 
and NFD normalization are equivalent. This can be shown by normalizing the 
example in either NFC or NFD, but using default separation in 
\texttt{write.profile}. To be precise, default separation means setting \texttt{sep = NULL} 
but that has not be added explicitly.

@@

\noindent Note that these different profiles can also be produced using the bash 
executable \texttt{writeprofile} (see \ref{implementations} for notes on how to 
install the bash executable). Exactly this example is also included in the help 
file of the executable.

@@

For quick solutions, it is also possible to leave out the Unicode information in
the profile skeleton by using the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simply situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

@@

\noindent This will lead to the following four files being written:

   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
         graphemes with added frequencies of occurence in the data. The lines
         in the file is re-ordered according to the order that resulted from the
         ordering specifications.

@@

   \item \textsc{size}\\
         order the lines in the profile by the size of the
         grapheme, largest first. Size is measured by number of Unicode
         characters after normalization as specified in the option
         \texttt{normalize}. For example, <é> has a size of 1 with
         \texttt{normalize = "NFC"}, but a size of 2 with
         \texttt{normalize = "NFD"}.

   \item \textsc{context}\\
         order the lines by whether they have any context
         specified (see next section). Lines with context coming first. Note
         that this only works when the option \texttt{regex = TRUE} is also chosen.

  \item \textsc{reverse}\\
        order the lines from bottom to top.

  \item \textsc{frequency}\\
         order the lines by the frequency with which they
         match in the specified strings before tokenization, least frequent
         coming first. This frequency of course depends crucially on the
         available strings, so it will lead to different orderings when applied
         to different data. Also note that this frequency is (necessarily)
         measured before graphemes are identified, so these ordering frequencies
         are not the same as the final frequencies shown in the output.
         Frequency of course also strongly differs on whether context is used
         for the matching through \texttt{regex = TRUE}.

@@

To document the process of graphemic tokenization, it is strongly encouraged to

@@

To refine a profile, it is also possible to add graphemes with contextual
specifications. An orthography profile can have columns called \texttt{Left} and
\texttt{Right} to specify the context in which the Grapheme is to be
separated.\footnote{The columns names \textttf{Left}, \textttf{Right} and
\textttf{Grapheme} are currently hard coded, so exactly these column names
should be used for these effects to take place. The position of the columns in
the profile is unimportant. So the column \textttf{Left} can occur anywhere.} We
are adding an extra line to the profile specifying that <s> is a grapheme when
it occurs after <mi>. These contextually specified graphemes are based on
regular expressions (so you can actually use regular expressions in the
description of the context). For these to be included in the graphemic parsing 
we have to specify the option \texttt{regex = TRUE}. This contextually specified 
grapheme should actually be handled first, so we could try \texttt{ordering = 
NULL}. However, we can also explicitly specifiy that rules with contextual 
information should be applied first by using \texttt{ordering = "context"}. That 
gives the right results for this toy example.

@@

\section{Profile skeleton with columns for editing}
\label{profile-editing})

When it is expected that context might be important for a profile, then the 
profile skeleton can be written to a file with columns prepared for the 
contextual specifications by using the option \texttt{editing = TRUE}.

>>>>

\noindent The version available through CRAN is the latest stable version.
To obtain the most recent bug-fixes and experimental additions, please use the
development version, which is available on
GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
version can be easily installed using the github-install helper software from the
\texttt{devtools} package.

<<eval=FALSE>>=
# download and install helper software
install.packages('devtools') 
# install the qlcData package from GitHub
devtools::install_github('cysouw/qlcData', build_vignettes = TRUE)
# load the software, so it can be used 
library(qlcData) 
@

@@

\noindent By specifying the splitting separator as the empty string
\texttt{sep~=~""}, it is possible to split the string into Unicode codepoints,
thus separating the combining diacritics. The idea behind this option
\texttt{sep} is that separating by a character allows for user-determined
separation. The most extreme choice here is the empty string \texttt{sep~=~""},
which is interpreted as separation everywhere. The other extreme is the default
setting \texttt{sep~=~NULL}, which means that the separation is not
user-defined, but relegated to the Unicode grapheme definitions.

@@

\noindent Some characters look identical, although they are encoded differently.
Unicode offers different ways of normalization (see
Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
using the option \texttt{normalize}. NFC-normalization turns everything into the
precomposed characters, while NFD-normalization separates everything into base
characters with combining diacritics. Splitting by codepoints (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.

@@

\noindent It is important to realize that for Unicode grapheme definitions, NFC
and NFD normalization are equivalent. This can be shown by normalizing the
example in either NFC or NFD, but using default separation in
\texttt{write.profile}. To be precise, default separation means setting
\texttt{sep~=~NULL}, but that has not be added explicitly.

@@

\noindent Note that these different profiles can also be produced using the bash
executable \texttt{writeprofile} (see \ref{available-implementations} for notes
on how to install the bash executable). Exactly this example is also included in
the help file of the executable.

@@

\noindent It is also possible to leave out the Unicode information in
the profile skeleton by using the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simply situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

@@

@@

   \item \textsc{size}\\
         This option orders the lines in the profile by the size of the
         grapheme, largest first. Size is measured by number of Unicode
         characters after normalization as specified in the option
         \texttt{normalize}. For example, <é> has a size of 1 with
         \texttt{normalize = "NFC"}, but a size of 2 with
         \texttt{normalize = "NFD"}.

   \item \textsc{context}\\ This option orders the lines by whether they have
           any context specified (see next section). Lines with context will
           then be used first. Note that this only works when the option
           \texttt{regex = TRUE} is also chosen (otherwise context
           specifications are not used).

   \item \textsc{reverse}\\ This option orders the lines from bottom to top.
         Reversing order can be useful because hand-written profiles tend to put
         general rules before specific rules, which mostly should be applied in
         reverse order.

  \item \textsc{frequency}\\
         This option orders the lines by the frequency with which they
         match in the specified strings before tokenization, least frequent
         coming first. This frequency of course depends crucially on the
         available strings, so it will lead to different orderings when applied
         to different data. Also note that this frequency is (necessarily)
         measured before graphemes are identified, so these ordering frequencies
         are not the same as the final frequencies shown in the output.
         Frequency of course also strongly differs on whether context is used
         for the matching through \texttt{regex = TRUE}.

@@

\noindent This will lead to the following four files being written. Crucially, a
new profile is produced with the re-ordered orthography profile. To reproduce
the tokenization, this re-ordered profile can be used with the option
\texttt{ordering~=~NULL}.

   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
         graphemes with added frequencies of occurence in the data. The lines
         in the file is re-ordered according to the order that resulted from the
         ordering specifications (see Section~\ref{rule-ordering}).

@@

To refine a profile, it is also possible to add graphemes with contextual
specifications. An orthography profile can have columns called \texttt{Left} and
\texttt{Right} to specify the context in which the grapheme is to be
separated.\footnote{The columns names \textttf{Left}, \textttf{Right} and
\textttf{Grapheme} are currently hard-coded, so exactly these column names
should be used for these effects to take place. The position of the columns in
the profile is unimportant. So the column \textttf{Left} can occur anywhere.}
For example, we are adding an extra line to the profile from above, resulting in
the profile shown in Table~\ref{tab:profile_skeleton2}. The extra line specifies
that <s> is a grapheme when it occurs after <mi>. Such contextually-specified
graphemes are based on regular expressions so you can also use regular
expressions in the description of the context. For such contextually specified
graphemes to be included in the graphemic parsing we have to specify the option
\texttt{regex = TRUE}. This contextually specified grapheme should actually be
handled first, so we could try \texttt{ordering = NULL}. However, we can also
explicitly specifiy that rules with contextual information should be applied
first by using \texttt{ordering = "context"}. That gives the right results for
this toy example.

@@

It is important to realize that with the option \texttt{regex = TRUE} all
content in the profile is treated as regular expressions, so the characters with
special meaning in regular expressions should be either omitted or escaped (by
putting a <\ \backslash\ > \textsc{reverse solidus} at \uni{005C} before the
character). Specifically, this concerns the following characters:

  \item[] <-> \textsc{hyphen-minus} at \uni{002D}
  \item[] <!> \textsc{exclamation mark} at \uni{0021}
  \item[] <?> \textsc{question mark} at \uni{003F}
  \item[] <.> \textsc{full stop} at \uni{002E}
  \item[] <(> \textsc{left parenthesis} at \uni{0028}
  \item[] <)> \textsc{right parenthesis} at \uni{0029}
  \item[] <[> \textsc{left square bracket} at \uni{005B}
  \item[] <]> \textsc{right square bracket} at \uni{005D}
  \item[] <\{> \textsc{left curly bracket} at \uni{007B}
  \item[] <\}> \textsc{right curly bracket} at \uni{007D}
  \item[] <|> \textsc{vertical line} at \uni{007C}
  \item[] <*> \textsc{asterisk} at \uni{002A}
  \item[] <\backslash> \textsc{reverse solidus} at \uni{005C}
  \item[] <ˆ> \textsc{circumflex accent} at \uni{005E}
  \item[] <+> \textsc{plus sign} at \uni{002B}
  \item[] <\$> \textsc{dollar sign} at \uni{0024}

\section{Profile skeleton with columns for editing}
\label{profile-editing}

When it is expected that context might be important for a profile, then the
profile skeleton can be created with columns prepared for the contextual
specifications. This is done by using the option \texttt{editing = TRUE} (cf.
Table~\ref{tab:profile_editing_1} for a toy profile of some Italian words).

<<eval=FALSE, tidy=FALSE>>=
example <- c('cane', 'cena', 'cine')
write.profile(example
              , file = "~/Desktop/profile_skeleton.txt"
              , editing = TRUE
              , info = FALSE
              )
@

<<results='asis', echo=FALSE>>=
example <- c('cane', 'cena', 'cine')
profile_editing <- write.profile(example, editing=T, info=F)
print(xtable(profile_editing
        , caption = 'Orthography profile with empty columns for editing contexts'
        , label = 'tab:profile_editing_1'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'htb'
  )
@

Besides the columns \texttt{Left}, \texttt{Grapheme}, and \texttt{Right} as
discussed in the previous sections, there are also columns \texttt{Class} and
\texttt{Replacement}. The column \texttt{Class} can be used to specify classes
of graphemes that can then be used in the contextual specification. The column
\texttt{Replacement} is just a copy of the column \textt{Grapheme}, providing a
skeleton to specify transliteration. The name of the columne
\texttt{Replacement} is not fixed -- there can actually be multiple columns with 
different kinds of transliterations in a single profile.

To achieve contextually determined replacements it is possible to use a regular
expression in the contextual column. For example, consider the edited toy
profile for Italian in Table~\ref{tab:profile_editing_2} (where <c> becomes /k/
except before <i,e>, then it becomes /tʃ/).

<<results='asis', echo=FALSE>>=
colnames(profile_editing)[5] <- 'IPA'
profile_editing[2,5] <- 'k'
profile_editing <- rbind(c('', 'c', '[ie]', '', 'tʃ'), profile_editing)
profile_editing <- profile_editing[c(1,2,6,3,4,5),]
print(xtable(profile_editing
        , caption = 'Orthography profile with regex as context'
        , label = 'tab:profile_editing_2'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'htb'
  )
@

To use this profile, you have to add the option \texttt{regex = TRUE}. Also note
that we have changed the name of the transliteration-column, so we have to tell
the tokenization process to use this column to transliterate. This is done by
adding the option \texttt{transliterate = "IPA"}.

<<eval=FALSE, tidy=FALSE>>=
# add a contextual grapheme, and then use the edited profile to tokenize
tokenize( example
         , profile = "~/Desktop/profile_skeleton.txt"
         , regex = TRUE
         , transliterate = "IPA"
        )$strings
@

<<echo=FALSE>>=
tokenize(example
  , profile = profile_editing
  , regex = TRUE
  , transliterate = 'IPA'
  )$strings
@

Another equivalent possibility is to use a column \texttt{Class} to specify a
class of graphemes, and then use this class in the specification of context.
This is useful to keep track of recurrent classes in larger profiles. You are
free to use any class-name you like, as long as it doesn't clash with the rest
of the profile. The example shown in Table~\ref{tab:profile_editing_3} should 
give the same result as obtained previously by using a regular expression.

<<results='asis', echo=FALSE>>=
profile_editing[1,3] <- 'Vfront'
profile_editing[5,4] <- 'Vfront'
profile_editing[6,4] <- 'Vfront'
print(xtable(profile_editing
        , caption = 'Orthography profile with Class as context'
        , label = 'tab:profile_editing_3'
        )
  , include.rownames = FALSE
  , booktabs = TRUE
  , size = 'scriptsize'
  , table.placement = 'htb'
  )
@

<<eval=FALSE, tidy=FALSE>>=
# add a class, and then use the edited profile to tokenize
tokenize( example
         , profile = "~/Desktop/profile_skeleton.txt"
         , regex = TRUE
         , transliterate = "IPA"
        )$strings
@

<<echo=FALSE>>=
tokenize(example
  , profile = profile_editing
  , regex = TRUE
  , transliterate = 'IPA'
  )$strings
@

\section{Formatting grapheme separation}
\label{formattingseparation}

In all examples above we have used the default formatting for grapheme
separation using space as a separator, which is obtained by the default setting
\texttt{sep = " "}. It is possible to specify any other separator here,
including the empty string, i.e. \texttt{sep = ""}. This will not show the
graphemic tokenization anymore (although it has of course been used in the
background).

<<eval=FALSE, tidy=FALSE>>=
# Use the empty string as separator
tokenize( example
         , profile = "~/Desktop/profile_skeleton.txt"
         , regex = TRUE
         , transliterate = "IPA"
         , sep = ""
        )$strings
@

<<echo=FALSE>>=
tokenize(example
  , profile = profile_editing
  , regex = TRUE
  , transliterate = 'IPA'
  , sep = ''
  )$strings
@

Normally, the separator specified should not occur in the data. If it does,
unexpected things might happen, so consider removing the chosen seperator from
your strings first. However, there is also an option \texttt{sep.replace} to
replace the separator with something else. When \code{sep.replace} is specified,
this mark is inserted in the string at those places where the separator occurs.
Typical usage in linguistics would be \texttt{sep = " ", sep.replace = "\#"} adding
spaces between graphemes and replacing spaces in the input string by hashes in
the output string.

<<tidy=FALSE>>=
# Replace separator in string to be tokenized
tokenize( "test test test"
         , sep = " "
         , sep.replace = "#"
        )$strings$tokenized
@

\section{Remaining issues}

Given a set of graphemes, there are at least two different methods to tokenize
strings. The first is called \texttt{method = "global"} here: this approach
takes the first grapheme in the profile, then matches this grapheme globally at
all places in the string, and then turns to the next string in the profile. To
us, this seems to be the most intuitive way for linguists to deal with graphemic
parsing. The other approach is called \texttt{method = "linear"} here: this
approach walks through the string from left to right. At the first character it
looks through all graphemes whether there is any match, and then walks further
to the end of the match and starts again. This approach is more alike to
finite-state rewrite rules (though note that it still works differently from
such rewrite rules, see Section~\ref{rule-ordering}).

In some special cases these two methods can lead to different results, but these
special situations are very unlikely to happen in natural language. The example
below shows that a string \texttt{'abc'} can be parsed differently in case of a
very special profile with a very special ordering of the graphemes.

<<tidy=FALSE>>=
# different parsing methods can lead to different results
# the global method first catches 'bc'
tokenize( "abc"
         , profile = c("bc","ab","a","c")
         , order = NULL
         , method = "global"
         )$strings
# the linear method catches the first grapheme, which is 'ab'
tokenize( "abc"
         , profile = c("bc","ab","a","c")
         , order = NULL
         , method = "linear"
         )$strings
@

The current R-implementation has a limitation when regular expressions are used. 
The problem is that overlapping matches are not captured. Everything works as 
expected without regular expressions, but there might be warnings/errors in case of 
\texttt{regex = TRUE}. However, just as in the previous issue, this should very 
rarely happen in natural language.

The problem can be exemplified by a sequence <bbbb> in which a 
grapheme <bb> should be matched. With the default \texttt{regex = FALSE} 
there are three possible matches, but with \texttt{regex = TRUE} only the first 
two or the last two are matched. The middle two are not matched because they 
overlap with the other matches. In the example below this leads to an error, 
because the second <bb> is not matched. We have not been able to produce a real 
example in any natural language in which this limitation might lead to an error.

<<tidy=FALSE>>=
# Everything perfekt without regular expressions
tokenize( "abbb"
        , profile = c("ab","bb")
        , order = NULL
        , regex = FALSE
        )$strings
# Matching with regular expressions does not catch overlap
tokenize( "abbb"
        , profile = c("ab","bb")
        , order = NULL
        , regex = TRUE
        )$strings
@
